---
title: "Linking Pipeline"
format: html
editor: visual
---

# Link ESM to Phone Use Data

This data processing pipeline links ESM data to the phone use data. The goal is a dataset that contains the ESM data per beep and then per beep also the phone use (per smartphone pp category) within the hour before the respective beep of the participants.

The pipeline performs the following steps

1.  Read the ESM dataset (ESD) which is to be enriched with the phone use data
2.  Read the category dataset, which identifies a category per specific phone app
3.  Read the phone use datasets (PUD) and merges them into one single dataset
4.  Per beep in the ESD
    1.  Filters the PUD for all app use events for that participant and for the allowed time window before the beep (1 hour)
    2.  Interpret the filtered PUD dataset, apply the app category of each app use
    3.  Compute usage time and app open frequency per app category
5.   Merge all processed PUD sub-datasets into one new one
6.  Join the ESM dataset witht he processed PUD

```{r}
library(dplyr)
library(readr)
```

```{r}
five_daily <- read_csv("data/cleaned_phone_use_noLOC/data/five_daily.csv")
```
